---
title: "Week 2 Exercises"
author: "Zhi Zhou"
date: "`r Sys.Date()`"
format: html
editor: visual
---

# Exercise 1

## Task 1

Set $u=F(x)$, i.e. $u=1-(x_0/x)^{\alpha}$, and solve for $x$:

$(x_0/x)^\alpha=1/u\rightarrow x_0/x=(1-u)^{1/\alpha}\rightarrow x=\frac{x_0}{(1-u)^{1/\alpha}}$

$\boxed{F^{-1}(u) = x_0\,(1-u)^{-1/\alpha} = \frac{x_0}{(1-u)^{1/\alpha}},\qquad 0<u<1}$

## Task 2

```{r}
library(tidyverse)
library(patchwork)

rpareto_inv <- function(n, x0, al) {
  # n i.i.d. uniform(0, 1)
  u <- runif(n)
  # apply the inverse cdf -> Pareto draws
  x0 * (1 - u)^(-1 / al)
}
```

## Task 3

```{r}
set.seed(20260812)
n <- 5000
x0 <- 2
alpha3 <- 3
x <- rpareto_inv(n, x0, alpha3)
```

## Task 4

```{r}
# theoretical density
dpareto <- function(x, x0, al) {
  ifelse(x >= x0, (al * x0^al) / (x^(al + 1)), 0)
}

# theoretical quantile
qpareto <- function(p, x0, al) {
  x0 * (1 - p) ^ (-1 / al)
}

# theoretical cdf
ppareto <- function(x, x0, al) {
  1 - (x0 / x)^al
}

# returns three panels side by side
#   a) histogram + theoretical density
#   b) Q-Q plot (sample quantiles vs theoretical quantile)
#   c)empirical cdf vs theoretical cdf
checks <- function(x, x0, al, tag) {
  # plotting positions used to evaluate the theoretical quantiles
  p <- (seq_along(x) - 0.5) / length(x)
  
  # one data frame for all panels
  d <- data.frame(x = x,
                  theoq = qpareto(p, x0, al),
                  sampleq = sort(x))
  
  # axis cap: show the bulk of the data, hide the extreme 0.1% tail
  xmax <- quantile(x, 0.999)
  
  # 
  xs <- seq(x0, max(x), length.out = 400)
  
  # a) histogram + theoretical density
  p1 <- ggplot(d, aes(x)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins = 60,
      fill = "steelblue",
      alpha = 0.6,
      color = "white"
    ) +
    geom_line(
      data = data.frame(x = xs, y = dpareto(xs, x0, al)),
      aes(x, y),
      color = "darkred",
      linewidth = 1
    ) +
    coord_cartesian(xlim = c(x0, xmax)) +
    labs(
      subtitle = paste("(a) histogram + density"),
      x = "x",
      y = "density"
    ) +
    theme_bw()
  
  # b) Q-Q plot
  p2 <- ggplot(d, aes(theoq, sampleq)) +
    geom_point(size = 0.6, color = "steelblue") +
    geom_abline(
      slope = 1,
      intercept = 0,
      color = "darkred",
      linewidth = 1
    ) +
    labs(subtitle = paste("(b) Q-Q plot"),
         x = "theorectical quantiles",
         y = "sample quantiles") +
    theme_bw()
  
  # empirical vs theoretical cdf
  p3 <- ggplot(d, aes(x)) +
    stat_ecdf(geom = "step",
              color = "steelblue",
              linewidth = 1) +
    geom_line(
      data = data.frame(x = xs, y = ppareto(xs, x0, al)),
      aes(x, y),
      color = "darkred",
      linewidth = 1
    ) +
    coord_cartesian(xlim = c(x0, xmax)) +
    labs(
      subtitle = paste("(c) empirical vs theoretical cdf"),
      x = "x",
      y = "F(x)"
    ) +
    theme_bw()
  
  # combine the three panels side by side
  wrap_plots(p1, p2, p3, ncol = 3) + plot_annotation(title = tag)
}

# produce and show the three diagnostic plots
plt_alpha3 <- checks(x, x0, al = alpha3, "alpha = 3")
plt_alpha3
```

## Task 5

```{r}
theomean3 <- alpha3 * x0 / (alpha3 - 1)
c(
  theoretical_mean = theomean3,
  sample_mean = mean(x),
  relative_diff = (mean(x) - theomean3) / theomean3
)
```

## Task 6

```{r}
alpha12 <- 1.2
set.seed(20260812)
y <- rpareto_inv(n, x0, alpha12)
theomean12 <- alpha12 * x0 / (alpha12 - 1)

plt_alpha12 <- checks(x, x0, al = alpha12, "alpha = 1.2")
plt_alpha12

set.seed(123)
reps <- 200
mean3 <- replicate(reps, mean(rpareto_inv(n, x0, alpha3)))
mean12 <- replicate(reps, mean(rpareto_inv(n, x0, alpha12)))
c(alpha3_sd_of_means = sd(mean3),
  alpha12_sd_of_means = sd(mean12))
summary(mean12)

stab <- data.frame(
  mean = c(mean3, mean12),
  alpha = factor(rep(c("alpha == 3", "alpha == 1.2"), each = reps))
)

plt_stab <- ggplot(stab, aes(mean)) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    alpha = 0.8,
    color = "white"
  ) +
  geom_vline(
    data = data.frame(theo = c(3, 12), alpha = factor(c(
      "alpha == 3", "alpha == 1.2"
    ))),
    aes(xintercept = theo),
    color = "darkred",
    linewidth = 1
  ) +
  facet_wrap( ~ alpha, scale = "free", labeller = label_parsed) +
  labs(x = "sample mean (n = 5000)", y = "count", subtitl = "sample mean over 200 replications") +
  theme_bw()
plt_stab
```

